\proglang{Stan}: A Probabilistic Programming Language

### 2014-02-18 src/docs/papers/stan-language/article.tex
updated stan paper

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.  As of version 2.1,
  \proglang{Stan} provides full Bayesian inference for continuous-variable
  models through Markov chain Monte Carlo methods such as
  the No-U-Turn sampler, an adaptive form of Hamiltonian Monte
  Carlo sampling.  Penalized maximum likelihood estimates are
  calculated using optimization methods such as the
  Broyden–Fletcher–Goldfarb–Shanno algorithm.

@@

These models turned out to be a challenge for
existing general purpose inference software.  A direct encoding in
\proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS} \citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to \proglang{JAGS}
using \proglang{C++} to overcome the cost of interpretation.  Although
this is much faster than looping in \proglang{JAGS}, it quickly became
clear that the root of the problem was the slow convergence of
conditional sampling when parameters were highly correlated 
in the posterior, as for time-series models and hierarchical models 
with interacted predictors.  We finally realized we needed a better sampler, 
not a more efficient implementation of Gibbs sampling.

@@

We thought we were home free at this point, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.  Realizing that we were doing redundant
calculations, we wrote a vectorized form of the normal distribution
for multiple variates with the same mean and scale, which sped things
up a bit.  Performance was lacking until we finally figured out how
to both eliminate redundant calculations and partially evaluate the
gradients using a combination of expression templates and
metaprogramming.

@@

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in \proglang{Stan} 2.1.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

@@

% \nocite{WainwrightJordan:2008}
\nocite{R:2013}
\clearpage
%\bibliographystyle{jss}
\bibliography{stan-paper}

>>>>

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.  As of version 2.2.0,
  \proglang{Stan} provides full Bayesian inference for continuous-variable
  models through Markov chain Monte Carlo methods such as
  the No-U-Turn sampler, an adaptive form of Hamiltonian Monte
  Carlo sampling.  Penalized maximum likelihood estimates are
  calculated using optimization methods such as the
  Broyden–Fletcher–Goldfarb–Shanno algorithm.

@@

These models turned out to be a challenge for existing general purpose
inference software.  A direct encoding in \proglang{BUGS}
\citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS}~\citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to
\proglang{JAGS} using \proglang{C++} to overcome the cost of
interpretation.  Although this is much faster than looping in
\proglang{JAGS}, it quickly became clear that the root of the problem
was the slow convergence of conditional sampling when parameters were
highly correlated in the posterior, as for time-series models and
hierarchical models with interacted predictors.  We finally realized
we needed a better sampler, not a more efficient implementation of
Gibbs sampling.

@@

We thought we were home free, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.  Realizing that we were doing redundant
calculations, we wrote a vectorized form of the normal distribution
for multiple variates with the same mean and scale, which sped things
up a bit.  Performance was lacking until we finally figured out how
to both eliminate redundant calculations and partially evaluate the
gradients using a combination of expression templates and
metaprogramming.

@@

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in \proglang{Stan} 2.2.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

@@

% \nocite{WainwrightJordan:2008}
\nocite{R:2013}
\clearpage
% implicit: \bibliographystyle{jss}
\bibliography{stan-paper}
